---
title: "Differential expression at pathway level using limma model on pathway enrichment scores to identify                differentially active pathways across groups/subgroups (pathway signature)" 
author: "Menna Arafat"
date: "2024-8-26"
output: html_document
---

## create output directory

```{r}
dir.create("output")
```

## load data
```{r}
ES= read.csv("data/single_sample_gsea_enrichment_scores.csv")
metadata= read.csv("data/metadata.csv")
```


## Apply limma for "Differential pathway analysis"

```{r}
model= model.matrix(~ 0+metadata$condition) %>% as.data.frame()
row.names(model)= metadata$sample
names(model)= gsub("metadata\\$condition", "", names(model))

library(limma)
fit <- lmFit(ES, model)
fit <- eBayes(fit)
res <- decideTests(fit, p.value=0.5) # remember to reset pvalue cut-off to 0.05
summary(res)
```

## DE pathways
### #differentially expressed pathways for timepoint 2 (coef=3)
```{r}
fit$coefficients %>% head()
tt <- topTable(fit, coef=3, n=Inf)
DEpwys <- rownames(tt)[tt$adj.P.Val <= 0.8] # remember to reset adjusted pvalue cut-off to 0.05
DEpwys_es = ES[DEpwys, ]

write.csv(tt, "output/differentially_expressed_pathways_limma_on_ES.csv", row.names = T)
write.csv(DEpwys_es , "output/DEpwys_ES.csv", row.names = T)
```

## GSVA scores have higher precision for larger gene sets

```{r}
gssizes <- geneSetSizes(ES)
plot(sqrt(gssizes), sqrt(fit$sigma), xlab="Sqrt(gene sets sizes)",
     ylab="Sqrt(standard deviation)", las=1, pch=".", cex=3)
```

## In such a setting, we can improve the analysis of differentially expressed pathways by using the limma-trend approach (Phipson et al. 2016)
## setting the trend parameter in the call to the eBayes() function to the vector of gene set sizes. 
##  Key Idea Behind Limma-Trend
##  In RNA-seq and pathway analysis, there can be a mean-variance trend—i.e.,
##  genes or pathways with higher expression tend to have larger variances.
##  The limma-trend approach takes this into account by allowing the prior variance estimate
##  to depend on the mean expression of the gene or pathway, rather than assuming a constant variance for all genes.

```{r}
# fit <- eBayes(fit, trend=gssizes)
# res <- decideTests(fit, p.value=0.01)
# summary(res)
```

## -------------------------------------------------------------------------------
## heatmap DE pathways

```{r}
# tt <- topTable(fit, coef=4, n=Inf)
# DEpwys <- rownames(tt)[tt$adj.P.Val <= 0.01]
# DEpwys_es = ES[DEpwys, ]
```

## for hierarchical clustering,to determine whether to use pearson (assume norrmally distributed data and linear relation) or spearman (assume non normally distributed data)

```{r}
DEpwys_es = ES[DEpwys, ]
row.names(DEpwys_es)= NULL
DEpwys_es= DEpwys_es %>% column_to_rownames("X")
shapiro.test(as.numeric(DEpwys_es[1, ]))  #' ## p-value <= 0.05 indicative of non normal distribution 
shapiro.test(as.numeric(DEpwys_es[, 1])) 

colorLegend <- c("darkolivegreen" , "steelblue","#ADFF2F" )
names(colorLegend) <- unique(metadata$condition)
sample.color.map = colorLegend[metadata$condition]
names(sample.color.map) <- metadata$sample

sampleClustering <- hclust(as.dist(1-cor(as.matrix(DEpwys_es), method="spearman")), #' ## pearson #' ## spearman
                           method="complete")
geneSetClustering <- hclust(as.dist(1-cor(t(DEpwys_es), method="spearman")),
                            method="complete")
palette <- colorRampPalette(c( "lightyellow2" ,"#DFC27D" , "#FCAA0FFF", "darkred"))(256)

png("output/heatmap_DE_pathsways.png", height=4100, width = 2650, res= 600)
heatmap(as.matrix(DEpwys_es), ColSideColors=sample.color.map, xlab="samples",
        ylab="", margins=c(2, 20),
        col = palette ,
        labRow=substr(gsub("_", " ", gsub("^KEGG_|^REACTOME_|^BIOCARTA_", "", rownames(DEpwys_es))), 1, 35),
        labCol="", scale="none", Colv=as.dendrogram(sampleClustering),
        Rowv=as.dendrogram(geneSetClustering)
         )
legend("right", inset = 0.001, names(colorLegend), fill=colorLegend, bg="white",
       cex = .4, title= "Phases") #' ##  #' ## x = .01, y =9
image.plot(zlim = range(DEpwys_es, na.rm = TRUE),
           legend.only = TRUE, 
           horizontal = TRUE, 
           legend.shrink = 0.3, 
           legend.width = 0.9,
           legend.mar= 4.5,
           col = palette , 
           legend.position = c(0.5, 0.5), 
           inset = c(-1.1, 0),
           legend.args = list(text = "Enrichment Score (ES)", side = 3, line = .5, cex = 0.6))
dev.off()
```

## Auxillary plots
## volcano plot for DE pathways

```{r}
tt <- topTable(fit, coef=2, n=Inf)
DEpwys <- rownames(tt)[tt$adj.P.Val <= 0.01] # remember to reset adjusted pvalue cut-off 


plot(tt$logFC, -log10(tt$P.Value), pch=16, cex=1, col=grey(0.75),
     main="", xlab="GSVA enrichment score difference", las=1,
     ylab=expression(-log[10]~~Raw~P-value))
abline(h=-log10(max(tt$P.Value[tt$adj.P.Val <= 0.01])),
       col=grey(0.5), lwd=1, lty=2)
points(tt$logFC[match(DEpwys, rownames(tt))],
       -log10(tt$P.Value[match(DEpwys, rownames(tt))]),
       pch=16, cex=1, col="darkred")
text(max(tt$logFC)*0.85, -log10(max(tt$P.Value[tt$adj.P.Val <= 0.01])),
     "1% FDR", pos=3)
```

## -------------------------------------------------------------------------------
## heatmap enrichment scores 

```{r}
row.names(ES)= NULL
ES= ES %>% column_to_rownames("X")

heatmap_data <- scale(as.matrix(ES[1:150,]))

p= Heatmap(
  matrix = as.matrix(heatmap_data ),
  name = "Enrichment Scores",
  col = colorRamp2(c(-3, 0, 3), c("#9370DB", "#F7F5F4","red" )),#' ## matlab::jet.colors(200),
  show_row_names = TRUE,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_column_names = T
  #' ## top_annotation  = ta
  #' ## column_title = ""
) 

print(p)
```

